OpenMPIR

### 2017-10-04 paper.tex
Addressed reviewers

\author{Kei Davis}
\affiliation{\institution{Los Alamos National Laboratory}}
\email{kei@lanl.gov}

@@

In this section we further discuss a significant challenge in HPC:
fragmentation of parallel programming models. There are many ways to write
parallel programs~\cite{openmp, qthreads, chapel, cilk, kokkos, legion, upc,
mpi}.  From libraries, to language extensions, to standalone languages, to
combinations of the these, the fragmentation of parallel programming models is a
problem that is only getting worse. This means that when writing tooling,
optimizations, or analyses, one much choose which model to target as well as
understand points of contention and interoperability between them. A common
difficulty among all of these is that reasoning about parallel programs is hard
and compilers can potentially help reduce this burden. This is exactly
analogous the problem LLVM solves: compiler optimizations are hard, so sharing
them is valuable. Tapir is a demonstration that the LLVM approach to
compilation of serial code can be effectively extended to help reason about
parallel code.

@@

To represent parallelism Tapir introduces the \code{detach}, \code{reattach}, and
\code{sync} instructions. The \code{detach} statement is syntactically similar to
a conditional branch in that it is a terminator instruction with with two blocks.
However, unlike the conditional branch which denotes a choice between successors, the \code{detach} instruction denotes that its two successors may run in parallel. Semantically,
the first successor of a \code{detach} instruction is referred to as the \textbf{detached block} and represents a task that can, but is not required to, run in parallel with the
\textbf{continuation block}, or the second successor to the detach statement. The \code{detach}
is analagous to a \code{fork} in most fork-join models.

The end of a task created by a \code{detach} instruction must end with a \code{reattach} instruction to the corresponding continuation block. The \code{reattach} instruction
is used to denote the end of a parallel task.
%
Finally, one may use the \code{sync} instruction to wait for any outstanding tasks created
by \code{detach} instructions within the current function to complete. The \code{sync}
instruction is analagous to a \code{join} in most fork-join frameworks.

@@

While we did implement the codegen for the \texttt{task} and
\texttt{taskwait} constructs, it's worth noting that even for these pragmas
the implementation is incomplete. Currently, any clauses modifying the behavior
of the pragmas is ignored. Probably the most common semantics this will change
is for the semantcis of variables, e.g.\ \texttt{shared} vs.\ \texttt{private}.
Surprisingly, this had little effect on the correctness of the entire Barcelona
OpenMP task suite. Indeed, it is likely that fixing this problem would
marginally increase performance by reducing the number of memory copies for
variables declared \texttt{private} in OpenMP code.  The fact that behaviour
wasn't changed significantly is worth revisiting, and we do so in
Section~\ref{Sec:Discussion}.

@@

There are other issues that had to be overcome to run full OpenMP programs
using Tapir. Generally, in addition the the \texttt{task} 
and \texttt{taskwait} pragmas, a program must contain pragmas to initialize OpenMP
%pragmas in order to run parallel code.
A standard pattern for building
task-parallel OpenMP programs is to insert a \texttt{parallel} pragma to start
the necessary hardware threads, followed immediately by a \texttt{single}
pragma to have only one thread continue on the specified statement. Then the
other threads will get work from spawned tasks instead of implicitly executing
the same code in a data parallel manner. For the purpose of this work we took
the shortcut of replacing these pragmas with no-ops. This worked well for all
examples except for one in which after the \texttt{parallel} and
\texttt{single} pragmas the top level task was called with an unnecessary
\texttt{task} pragma.  Because the wait was implicit and unhandled by our
implementation, the simple fix was to remove the \texttt{task} pragma. This was
the only change required to the source code. Our temporary no-op shortcut, of
course, doesn't follow the OpenMP specification, and should be addressed by
future work. Section~\ref{Sec:Future} discusses how how the 
implementation can be improved.

@@

Upon entering the \texttt{else} branch, the code immediately detaches the basic
block corresponding to the call to \texttt{fib(n-1)}, labeled
\texttt{det.achd}. The continuation for the first call also immediately
detaches the basic block corresponding to the call of \texttt{fib(n-2)}.
Finally, the continuation for the second call immediately calls sync. This is
a nice example of code that will be optimized by Tapir. The LLVM code
shown in \figref{tapir-example} is not optimized. With optimizations turned
on Tapir will replace a \texttt{detach} of a function call followed
immediately by \texttt{sync} with a simple function call, resulting in faster
code.

@@

As mentioned in the Section~\ref{Sec:Implementation}, the gaps in the implementation
changed program behavior in a couple cases. In the case of \texttt{FFT}, we were
forced to remove an unnecessary \texttt{task} pragma, as the current implementation
doesn't insert the implicit barrier at the end of a \texttt{parallel} region. This 
was a one line fix in the benchmark, and would be fixed properly by handling
the \texttt{parallel} and \texttt{single} pragmas correctly in our implementation.

@@

In this section we speculate \emph{why} the performance varies in
the ways it does. While we have yet to identify every discrepancy in
performance, we hope to get some ideas for why the performance of our
implementation is generally better and where each of the implementations is
spending the bulk of the execution time. Our primary tool for this task is performance 
counters. It's worth noting that some information is difficult to infer from 
performance counters, such as sources of contention, reasons for memory locality,
etc. Still, they do give us some insight into performance bottlenecks.

Before we begin understanding why, it's worth noting again that the scope of the 
work that this paper is responsible for is only the front end. All performance
gains attributable to the existing Tapir back end and Cilk runtime represent prior work.
We can break reasons for performance discrepancies into a few categories: 
\begin{itemize}
\item Front-end code generation;
\item IR Optimizations; and
\item Runtime efficiency.
\end{itemize}
While we would like to distinguish between these, it is difficult to do so with
only performance counter information. A more thorough investigation would
require deep knowledge of each of the front ends and runtimes. While the
importance of optimizations of parallel code is clear, we hypothesize that most
of the disparity present in these results is more a function of runtime
implementation differences. As justification for this hypothesis, consider
the \texttt{Fibonacci} benchmark. There are no memory accesses to the heap, 
and the code leaves little room for optimization (algorithmic rewrites
aside), so it is effectively a benchmark of how fast a
runtime can execute fork-join parallelism.

@@

There are two cases we examine using the profiler. The first is the
case when there is a large difference in run times between implementations. For 
these our hypothesis is that the slower implementations
are for some reason spending more time in the runtime. As mentioned, 
it is difficult to understand exactly why an implementation is spending more 
time in the runtime without further investigation, but it is a useful 
sanity check nonetheless. For this case, we turn to the \texttt{Fib} benchmark.
Indeed, we get roughly the following breakdowns of work-to-overhead ratio
according to the Linux \texttt{perf} tool.
\begin{itemize}
\item \texttt{tapir: ~30\% runtime overhead, ~50\% work, ~20\% other}
\item \texttt{clang: ~85\% runtime overhead, ~5\% work, ~10\% other}
\item \texttt{icc: ~85\% runtime overhead, ~5\% work, ~10\% other}
\item \texttt{gcc: ~100\% runtime overhead, ~0\% work, ~0\% other}
\end{itemize}
This supports our hypothesis that most of the slowdown incurred is overhead in
the runtime. For Tapir the overhead is relatively small given the
size of the work chunks, for \texttt{clang} and \texttt{icc} it is
significantly larger, and for \texttt{gcc} it is overwhelming.

@@

One interesting aspect of this work is that it compiles a language extension
intended for one runtime (OpenMP) to use another (the Cilk runtime). While in
many ways this is a weakness of this work, it does open the possibility of
utilizing this in a more general way to address many significant challenges
such as performance portability. Having a parallel-aware IR enables the
possibility of multiple high-level parallel languages and multiple parallel
runtime targets in the same way that LLVM does for serial code. While the
OpenMP specification's requirement to have access to runtime routines
complicates things, for many programming models it seems at least conceivable
to map them onto multiple different runtimes. This opens the possibility of
mixing programming models at compile time by targeting the same runtime. For
example, if both OpenMP and Cilk targeted Tapir instructions, one could have a
large application where some parts were written using Cilk, while other used
OpenMP, and the result would not suffer the incompatibility problems often
seen today.

The surprising fact that program behavior wasn't changed significantly is
worth discussing here. We surmise that it is likely due to the
semantics of variables borrowed from the Cilk codegen and their relation to the OpenMP
specification of semantics. In particular, \emph{OpenMP doesn't specify a
memory model for shared variables}. This results in the Cilk style of only
writing to a shared variable at the end of a parallel section as a perfectly
valid implementation of OpenMP's specification. In contrast, existing OpenMP
implementations often write to shared variables on every write. This is a
potential for performance discrepancy because of differing implementation-defined
behavior and we will return to it when discussing performance on the
Barcelona OpenMP task suite. There is of course the fact that technically, by
copying the value of variables back to the surrounding context even for
\texttt{private} variables, one isn't following the specification. We surmise
that in practice this hasn't had an effect because
\texttt{private} variables are generally used for performance rather than
their semantic properties.

@@

A related approach to composing programming models is to keep the same code
generation approach, but replace calls into built in runtimes with calls into a
shared runtime. This is the approach taken by Lithe, where Pan et
al.\ compose programs using both GNU OpenMP and Intel TBB~\cite{lithe}. This approach succeeds
in composing different programming models at run-time, but doesn't enable the
static analyses that enabled by Tapir.

@@

@@

This work was supported by the Director, Office of Advanced Scientific
Computing Research, Office of Science, of the United States Department
of Energy, under the guidance of Dr. Sonia Sachs. Los Alamos National
Laboratory is operated by Los Alamos National Security LLC for the U.S.
Department of Energy under contract DE-AC52-06NA25396.

This research was supported by US National Science Foundation under CRII
ACI 1565338, CNS 1527076, and CNS 1217948. Los Alamos National
Laboratory is operated by Los Alamos National Security LLC for the US
Department of Energy under contract DE-AC52-06NA25396.

Work supported by the Advanced Simulation and Computing program of the
U.S. Department of Energy's NNSA\@.  Los Alamos National Laboratory is
managed and operated by Los Alamos National Security, LLC (LANS), under
contract number DE-AC52-06NA25396 for the Department of Energy's
National Nuclear Security Administration (NNSA).

>>>>

@@

In this section we further discuss a significant challenge in HPC:
fragmentation of parallel programming models. There are many ways to write
parallel programs~\cite{openmp, qthreads, chapel, cilk, kokkos, legion, upc,
mpi}.  From libraries, to language extensions, to standalone languages, to
combinations of the these, the fragmentation of parallel programming models is a
problem that is only getting worse. This means that when writing tooling,
optimizations, or analyses, one much choose which model to target as well as
understand points of contention and interoperability between them. A common
difficulty among all of these is that reasoning about parallel programs is hard
and compilers can potentially help reduce this burden. This is exactly
analogous to the problem LLVM solves: compiler optimizations are hard, so
sharing them is valuable. Tapir is a demonstration that the LLVM approach to
compilation of serial code can be effectively extended to help reason about
parallel code.

@@

To represent parallelism Tapir introduces the \code{detach}, \code{reattach}, and
\code{sync} instructions. The \code{detach} statement is syntactically similar to
a conditional branch in that it is a terminator instruction with two blocks.
However, unlike the conditional branch which denotes a choice between successors, the \code{detach} instruction denotes that its two successors may run in parallel. Semantically,
the first successor of a \code{detach} instruction is referred to as the \textbf{detached block} and represents a task that can, but is not required to, run in parallel with the
\textbf{continuation block}, or the second successor to the detach statement. The \code{detach}
is analogous to a \code{fork} in most fork-join models.

The end of a task created by a \code{detach} instruction must end with a \code{reattach} instruction to the corresponding continuation block. The \code{reattach} instruction
is used to denote the end of a parallel task.
%
Finally, one may use the \code{sync} instruction to wait for any outstanding tasks created
by \code{detach} instructions within the current function to complete. The \code{sync}
instruction is analogous to a \code{join} in most fork-join frameworks.

@@

While we did implement the codegen for the \texttt{task} and
\texttt{taskwait} constructs, it's worth noting that even for these pragmas
the implementation is incomplete. Currently, any clauses modifying the behavior
of the pragmas is ignored. Probably the most common semantics this will change
is for the semantics of variables, e.g.\ \texttt{shared} vs.\ \texttt{private}.
Surprisingly, this had little effect on the correctness of the entire Barcelona
OpenMP task suite. Indeed, it is likely that fixing this problem would
marginally increase performance by reducing the number of memory copies for
variables declared \texttt{private} in OpenMP code.  The fact that behaviour
wasn't changed significantly is worth revisiting, and we do so in
Section~\ref{Sec:Discussion}.

@@

There are other issues that had to be overcome to run full OpenMP programs
using Tapir. Generally, in addition the the \texttt{task} 
and \texttt{taskwait} pragmas, a program must contain pragmas to initialize OpenMP.
A standard pattern for building task-parallel OpenMP programs is to insert a
\texttt{parallel} pragma to start the necessary hardware threads, followed
immediately by a \texttt{single} pragma to have only one thread continue on the
specified statement. Then the other threads will get work from spawned tasks
instead of implicitly executing the same code in a data parallel manner. For
the purpose of this work we took the shortcut of replacing these pragmas with
no-ops. This worked well for all examples except for one in which after the
\texttt{parallel} and \texttt{single} pragmas the top level task was called
with an unnecessary \texttt{task} pragma.  Because the wait was implicit and
unhandled by our implementation, the simple fix was to remove the \texttt{task}
pragma. This was the only change required to the source code. Our temporary
no-op shortcut, of course, doesn't follow the OpenMP specification, and should
be addressed by future work. Section~\ref{Sec:Future} discusses how how the
implementation can be improved.

@@

Upon entering the \texttt{else} branch, the code immediately detaches the basic
block corresponding to the call to \texttt{fib(n-1)}, labeled
\texttt{det.achd}. The continuation for the first call also immediately
detaches the basic block corresponding to the call of \texttt{fib(n-2)}.
Finally, the continuation for the second call immediately calls sync. This is
a nice example of code that will be optimized by Tapir. The LLVM code
shown in \figref{tapir-example} is not optimized. With optimizations turned
on, Tapir will replace a \texttt{detach} of a function call followed
immediately by \texttt{sync} with a simple function call, resulting in faster
code.

@@

As mentioned in the Section~\ref{Sec:Implementation}, the gaps in the implementation
changed program behavior in a couple of cases. In the case of \texttt{FFT}, we were
forced to remove an unnecessary \texttt{task} pragma, as the current implementation
doesn't insert the implicit barrier at the end of a \texttt{parallel} region. This 
was a one line fix in the benchmark, and would be fixed properly by handling
the \texttt{parallel} and \texttt{single} pragmas correctly in our implementation.

@@

In this section we investigate \emph{why} performance varies in the ways it
does. While we have yet to identify every discrepancy in performance, we hope
to get some ideas for why the performance of our implementation is generally
better and where each of the implementations is spending the bulk of the
execution time. Our primary tool for this task is profiling. It's worth noting
that lots of information is difficult to infer from profiling, such as sources
of contention, reasons for memory locality, etc.

It is worth noting again that the scope of the work that this paper is
responsible for is only the front end. All performance gains attributable to
the existing Tapir back end and Cilk runtime represent prior work.  We can
break reasons for performance discrepancies into a few categories:
\begin{itemize}
\item Front-end code generation;
\item IR Optimizations; and
\item Runtime efficiency.
\end{itemize}
While we would like to distinguish between these, it is difficult to do so with
only profiling information. A more thorough investigation would require deep
knowledge of each of the front ends and runtimes. While the importance of
optimizations of parallel code is clear, we hypothesize that most of the
disparity present in these results is more a function of runtime implementation
differences. As justification for this hypothesis, consider the
\texttt{Fibonacci} benchmark. There are no memory accesses to the heap, and the
code leaves little room for optimization (algorithmic rewrites aside), so it is
effectively a benchmark of how fast a runtime can execute fork-join
parallelism.

@@

There are two cases we examine using the profiler. The first is the
case when there is a large difference in run times between implementations. For 
these our hypothesis is that the slower implementations
are for some reason spending more time in the runtime. As mentioned, 
it is difficult to understand exactly why an implementation is spending more 
time in the runtime without further investigation, but it is a useful 
sanity check nonetheless. For this case, we turn to the \texttt{Fib} benchmark.
Indeed, we get roughly the following breakdowns of work-to-overhead ratio
according to the Linux \texttt{perf} tool.
\begin{itemize}
\item \texttt{tapir: ~30\% runtime overhead, ~50\% work, ~20\% other}
\item \texttt{clang: ~85\% runtime overhead, ~5\% work, ~10\% other}
\item \texttt{icc: ~85\% runtime overhead, ~5\% work, ~10\% other}
\item \texttt{gcc: ~100\% runtime overhead, ~0\% work, ~0\% other}
\end{itemize}
This supports our hypothesis that most of the slowdown incurred is overhead in
the runtime. For Tapir the overhead is relatively small given the size of the
work chunks, for \texttt{clang} and \texttt{icc} it is significantly larger,
and for \texttt{gcc} it is overwhelming. It's worth noting that \texttt{clang} 
and \texttt{icc} have effectively the same runtime, so the fact that their 
overheads are similar is unsurprising.

@@

One interesting aspect of this work is that it compiles a language extension
intended for one runtime (OpenMP) to use another (the Cilk runtime). While in
many ways this is a weakness of this work, it does open the possibility of
utilizing this in a more general way to address many significant challenges
such as performance portability. Having a parallel-aware IR enables the
possibility of multiple high-level parallel languages and multiple parallel
runtime targets in the same way that LLVM does for serial code. While the
OpenMP specification's requirement to have access to runtime routines
complicates things, for many programming models it seems at least conceivable
to map them onto multiple different runtimes. This opens the possibility of
mixing programming models at compile time by targeting the same runtime. For
example, if both OpenMP and Cilk targeted Tapir instructions, one could have a
large application where some parts were written using Cilk, while others used
OpenMP, and the result would not suffer the incompatibility problems often
seen today.

The surprising fact that program behavior wasn't changed significantly is worth
discussing. We surmise that it is likely due to the semantics of variables
borrowed from the Cilk codegen and their relation to the OpenMP specification
of semantics. In particular, \emph{OpenMP specifies a very relaxed memory model
for shared variables}. This results in the Cilk implementation of only writing
to a shared variable at the end of a parallel section being a perfectly valid
implementation of OpenMP's specification. In contrast, existing OpenMP
implementations often write to shared variables on every write. This is a
potential for performance discrepancy because of differing
implementation-defined behavior and we will return to it when discussing
performance on the Barcelona OpenMP task suite. There is of course the fact
that technically, by copying the value of variables back to the surrounding
context even for \texttt{private} variables, one isn't following the
specification. We surmise that in practice this hasn't had an effect because
\texttt{private} variables are generally used for performance rather than their
semantic properties.

@@

A related approach to composing programming models is to keep the same code
generation approach, but replace calls into built in runtimes with calls into a
shared runtime. This is the approach taken by Lithe, where Pan et
al.\ compose programs using both GNU OpenMP and Intel TBB~\cite{lithe}. This approach succeeds
in composing different programming models at run-time, but doesn't enable the
static analyses that are enabled by Tapir.

@@

Many thanks to Kei Davis for his excellent editing.

@@

This work was supported by the Director, Office of Advanced Scientific
Computing Research, Office of Science, of the United States Department of
Energy, under the guidance of Dr. Sonia Sachs. This research was supported by
US National Science Foundation under CRII ACI 1565338, CNS 1527076, and CNS
1217948.  Work supported by the Advanced Simulation and Computing program of
the U.S. Department of Energy's NNSA\@. Los Alamos National Laboratory is
managed and operated by Los Alamos National Security, LLC (LANS), under
contract number DE-AC52-06NA25396 for the Department of Energy's National
Nuclear Security Administration (NNSA).
